Stateless Model Checking with Data-Race Preemption Points

### 2015-11-16 pldi-16/di.tex
implement pt2 of joshua comments

With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
Note that even though a data race may involve two different instructions, we add new state spaces with only one new PP at a time.
Rather than adding a single state space configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

@@

%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
Note that under a {\em pure happens-before} analysis,
these accesses are not considered concurrent % at all
because of {\tt malloc}'s internal locking events,
and would not result in such false positives.
However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
so in our context it is more appropriate to use the
{\em limited happens-before} relation in a hybrid approach with lockset tracking.
%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,

>>>>

With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
Rather than adding a single large state space, %configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

@@

%%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
%Note that under a {\em pure happens-before} analysis,
%these accesses are not considered concurrent % at all
%because of {\tt malloc}'s internal locking events,
%and would not result in such false positives.
%However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
%so in our context it is more appropriate to use the
%{\em limited happens-before} relation in a hybrid approach with lockset tracking.
%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,

### 2015-11-16 pldi-16/evaluation.tex
implement pt2 of joshua comments

\subsection{Test Suite}
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtual memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
Though not ``real world'' programs, both projects are quite complex:
the P2s average 1807 lines of code,%C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/daniel-deniz*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718.

@@

\newcommand\mxtest{\texttt{mx\_test}} % TODO: find a way to change name to mutex_test without overflowing lines in the next paragraph
\newcommand\tej{\texttt{join\_test}}
\newcommand\bct{\texttt{bcast\_test}}
\newcommand\paraguay{\texttt{signal\_test}}
\newcommand\paradise{\texttt{sem\_test}}
\newcommand\rwldgr{\texttt{rwlock\_test}}
We tested P2s with 6 multithreaded programs:
% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
% each tailored to exercise a different part of the P2 project
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for cvars,
\paradise~for semaphores,
and \rwldgr~for r/w locks.
For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
\newcommand\prisema{\texttt{sched\_test}}
\newcommand\waitsimple{\texttt{wait\_test}}
\newcommand\alarmsimul{\texttt{alarm\_test}}
% TODO CAMREADY: Add another test case
We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented, so each test could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
}
The source code of all 9 tests is available at
{\em [submitted as supplementary material]}.
For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
In total, the test suite comprises 629 unique state spaces,
at least 165 of which contain bugs.

@@

{\bf Finding the same bugs faster.}
To show that Iterative Deepening is effective even for MC domains without data races, such as message-passing distributed systems,
we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
(i.e., ignoring all data-race candidates)\footnote{
Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
we ran these tests for 2.5 hours on 4 CPUs each.
Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
%
The line QS-no-DR-PPs represents this experiment.
Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
QS-no-DR-PPs finds many more of the bugs much sooner.
From this we conclude that for smaller arbitrary CPU budgets,
Iterative Deepening is likely to find bugs SSS-MC will miss,
and for more ambitiously-sized tests,
programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
%We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
%Iterative Deepening is likely to find bugs SSS-MC will miss.
%Moreover, it is also easy to imagine scaling up the size of each test case to test ,
%using more threads or longer sequences of API calls.
%We hope that is compelling even to users willing to spend many CPU-hours on testing.
%These results show explicitly that for arbitrary CPU-time budgets

@@

\begin{table*}[t]
	\begin{center}
		%\small
	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
		% TODO: Make the "bugs" left half of this graph into a bar graph.
		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Untested} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf DR PPs} & {\bf recycle DRs} & {\bf DR PPs} & {\bf DR bugs} \\
		\hline
		{\bct		} & 79	& 7	& 11	& 4	&	&	& 65	&	& 4	\\
		{\tej		} & 79	& 14 	& 25	& 11	&	&	& 471	&	& 6	\\
		{\mxtest	} & 79	& 1	& 10	& 9	&	&	& 7	&	& 1	\\
		{\paradise	} & 79	& 10	& 17	& 7	&	&	& 268	&	& 3	\\
		{\paraguay	} & 79	& 4	& 12	& 9	&	&	& 188	&	& 0	\\
		{\rwldgr	} & 79	& 26	& 30	& 3	&	&	& 157	&	& 3	\\
		\hline
		{\prisema	} & 59	&	&	&	&	&	& 0	&	&	\\
		{\alarmsimul	} & 44	&	&	&	&	&	& 37	&	&	\\
		{\waitsimple	} & 52	&	&	&	&	&	& 57	&	&	\\
		\hline
		{\bf Total}	& 629	&	&	&	&	&	& 1250	&	&
	\end{tabular}
	\end{center}
	\caption{Summary of number and types of bugs found by each test case.
		SSS-MC is the control; QS is \quicksand.
		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}}).
	``Nondeterministic DRs'' counts among the DR-only bugs how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
		TODO: more explanation}
	\label{tab:drbugs}
\end{table*}

@@

%Figure~\ref{fig:bargraph}
The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test case.
% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
% (as well as on the mutex API boundaries).
% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
even compared to the other test cases\footnote{
% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
	The one bug SSS-MC found was in a fully-assembly lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
%Though for many applications of MC, assuming a correct lock implementation is sufficient,
Though it suffices for many applications of MC to assume correctly-implemented locks,
we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.

{\bf Partial verification.}
When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
Beyond using the state space estimator to guess the percent coverage,
we know of no approach to quantify the probability
that a bug would be exposed by an untested interleaving.
\quicksand~does not attempt any such quantitative guarantee,
but can at least report which subsets of PPs produced state spaces that did complete in time.

The SSS-MC control experiment timed out in 10 hours on 77 tests.
Among these, \quicksand~found data-race bugs in 15.
For the other 62, we counted the number of state spaces \quicksand~was able to complete and the number of unique PPs among them.
In three cases, \quicksand~also failed to complete anything; beyond these,
between 1 and 253 state spaces were completed, with a mean of 58.9 and median of 29.
Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
These completions guarantee that, if the test case could expose a bug,
% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.

@@

{\bf Suppressing ``malloc-recycle'' false positives.}
In \sect{\ref{sec:recycle}} we showed the soundness of suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test case.
In total, 1250 data-races fit the malloc-recycle pattern across all tests,
only 69 of which were observed to avoid the re-allocation in an alternate interleaving.
Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 1181 other state spaces.

Among the 69 true data-races which initially fit the malloc-recycle pattern,
none of them exposed a new bug when used as a PP.
This shows that for other data-race tools,
suppressing malloc-recycle candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, \quicksand~was able to correctly identify the 69 violations of that heuristic among 30 distinct tests,
and fall back to classifying them with DPOR,
which we consider a much stronger verification.

{\bf Full verification.}
For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU-time.
Among these, 36 tests contained no data-race candidates whatsoever, 
so the same verification could be provided by SSS-MC.
We plot these for comparison,
according to the time taken by that test's SSS-MC control experiment,.
When a test can be totally verified without any data-race PPs,
\quicksand~will be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
Of course, using data-race PPs increases our capacity for total verification by 4.25x.

@@

\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}
Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
We classified each data-race candidate according to whether \landslide~reported them during the first interleaving,
before any backtracking or preempting:
if so, they were {\em single-pass data races}; otherwise, {\em nondeterministic}.

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{nondet-drs-1-v2.pdf}
	\caption{Some data-race candidates may not be identified during a single program execution.
		Using nondeterministic data races as PPs,
		\quicksand~found 189\% as many bugs compared to the approach of prior work \cite{portend}.
	}
	\label{fig:dr-falsenegs}
\end{figure}
Figure~\ref{fig:dr-falsenegs} compares the types of data-race candidates necessary to expose each data-race bug in our test suite.
The first series represents the bugs found using PPs from single-pass data-race candidates,
% TODO: re read racefuzzer/portend paper to confirm this is true.
i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
The second series shows all the data-race bugs \quicksand~found,
which includes the former type as well as new bugs involving nondeterministic data-races.
In total, among \quicksand's 68 total data-race bugs, just 36 could be found with single-pass data-race candidates alone.
%In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.

At first, it might seem unfair to compare the data race bugs \quicksand~finds (10 CPU-hours)
against the candidates found by a prior work's single execution (less than a minute).
We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
are not intended to discover new candidates under subsequent runs. %before discovering any initial candidates.
Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
\cite{chess-icb,gambit}.
This result suggests that
other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use a similar feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
%i.e., discovering transitively-reachable data races while testing initial ones.
%although that would still not simultaneously be able to provide total verifications.
% TODO: read gambit paper

@@

{\bf Finer-grained PP subsets.}
% TODO: update these numbers
\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
However, in 2 tests, no more than the minimal state space could be verified,
and in 3 other tests, not even that much.
Larger state spaces often result from finer-grained locking,
which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
Such programs may require even more rigorous verification than a program with a single global lock.
%Hence these corner cases are important to consider for future work.
While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
making partial verification of such large tests possible.
%enabling partial verification of such large tests. % if need space
%Because our current implementation does not avoid repeated interleavings across state spaces,
%as discussed above,
%we were limited to a small number of very basic PP subsets to statically seed the exploration

@@

Likewise, for test cases where full verification is not computationally feasible,
the state spaces associated with many data-race candidates will time out.
We cannot guarantee that those races are false positives or benign, even though no bug was found.
% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test case.
For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
which could be combined with Iterative Deepening in future work.
% Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

>>>>

\subsection{Test Suite}
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtual memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
Though not ``real world'' programs, both projects are quite complex:
the P2s average 1807 lines of code, %C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/daniel-deniz*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718.

@@

\newcommand\mxtest{\texttt{mx\_test}} % TODO: find a way to change name to mutex_test without overflowing lines in the next paragraph
\newcommand\tej{\texttt{join\_test}}
\newcommand\bct{\texttt{bcast\_test}}
\newcommand\paraguay{\texttt{signal\_test}}
\newcommand\paradise{\texttt{sem\_test}}
\newcommand\rwldgr{\texttt{rwlock\_test}}
We tested P2s with 6 multithreaded programs:
% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
% each tailored to exercise a different part of the P2 project
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for cvars,
\paradise~for semaphores,
and \rwldgr~for r/w locks.
For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
\newcommand\prisema{\texttt{sched\_test}}
\newcommand\waitsimple{\texttt{wait\_test}}
\newcommand\alarmsimul{\texttt{alarm\_test}}
We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented, so each test program could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
}
The source code of all 9 test cases is available at
{\em [submitted as supplementary material]}.
For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
We ran all tests on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
In total, our evaluation comprises 629 tests (pairs of a test program and a Pintos or P2),
at least 163 of which could expose bugs.

@@

{\bf Finding the same bugs faster.}
To show that Iterative Deepening is effective even for MC domains without data races, such as message-passing distributed systems,
we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
(i.e., ignoring all data-race candidates).
%\footnote{
%Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
%we ran these tests for 2.5 hours on 4 CPUs each.
%Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
%
The line QS-no-DR-PPs represents this experiment.
Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
QS-no-DR-PPs finds many more of the bugs much sooner.
Hence, for smaller arbitrary CPU budgets,
Iterative Deepening is likely to find bugs SSS-MC will miss.
%and for more ambitiously-sized tests,
%programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
%We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
%Iterative Deepening is likely to find bugs SSS-MC will miss.
%Moreover, it is also easy to imagine scaling up the size of each test case to test ,
%using more threads or longer sequences of API calls.
%We hope that is compelling even to users willing to spend many CPU-hours on testing.
%These results show explicitly that for arbitrary CPU-time budgets

@@

\begin{table*}[t]
	\begin{center}
		%\small
	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
		% TODO: Make the "bugs" left half of this graph into a bar graph.
		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Untested} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf DR PPs} & {\bf recycle DRs} & {\bf DR PPs} & {\bf DR bugs} \\
		\hline
		{\bct		} & 79	& 7	& 11	& 4	&	&	& 65	&	& 4	\\
		{\tej		} & 79	& 14 	& 25	& 11	&	&	& 471	&	& 6	\\
		{\mxtest	} & 79	& 1	& 10	& 9	&	&	& 7	&	& 1	\\
		{\paradise	} & 79	& 10	& 17	& 7	&	&	& 268	&	& 3	\\
		{\paraguay	} & 79	& 4	& 12	& 9	&	&	& 188	&	& 0	\\
		{\rwldgr	} & 79	& 26	& 30	& 3	&	&	& 157	&	& 3	\\
		\hline
		{\prisema	} & 59	&	&	&	&	&	& 0	&	&	\\
		{\alarmsimul	} & 44	&	&	&	&	&	& 37	&	&	\\
		{\waitsimple	} & 52	&	&	&	&	&	& 57	&	&	\\
		\hline
		{\bf Total}	& 629	&	&	&	&	&	& 1250	&	&
	\end{tabular}
	\end{center}
	\caption{Summary of number and types of bugs found by each test program.
		SSS-MC is the control; QS is \quicksand.
		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}}).
	``Nondeterministic DRs'' counts among the DR-only bugs how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
		TODO: more explanation}
	\label{tab:drbugs}
\end{table*}

@@

%Figure~\ref{fig:bargraph}
The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test program.
% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
% (as well as on the mutex API boundaries).
% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
even compared to the other test programs\footnote{
	The one bug SSS-MC found was in a fully-assembly lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
%Though for many applications of MC, assuming a correct lock implementation is sufficient,
Though it suffices for many applications of MC to assume correctly-implemented locks,
we consider this strong evidence that low-level synchronization code must be verified with data-race PPs.

% TODO: maybe the whole story about partial verification can be cut
{\bf Partial verification.}
When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
Beyond using the state space estimator to guess the percent coverage,
we know of no approach to quantify the probability
that a bug would be exposed by an untested interleaving.
\quicksand~does not attempt any such quantitative guarantee,
but can at least report which subsets of PPs produced state spaces that did complete in time.

On 77 tests, the SSS-MC control experiment timed out in 10 hours with no bugs found.
Among these, \quicksand~found data-race bugs in 15.
For the other 62, we counted the number of state spaces \quicksand~was able to complete and the number of unique PPs among them.
In three cases, \quicksand~also failed to complete anything; beyond these,
between 1 and 253 state spaces were completed, with a mean of 58.9 and median of 29.
Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
These completions guarantee that, if the test program could expose a bug,
% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.

@@

{\bf Suppressing ``malloc-recycle'' false positives.}
In \sect{\ref{sec:recycle}} we showed the soundness of suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
In total, 1250 data-races fit the malloc-recycle pattern across all tests,
only 69 of which were observed to avoid the re-allocation in an alternate interleaving.
Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 1181 other state spaces.

Among the 69 true data-races which initially fit the malloc-recycle pattern,
none of them exposed a new bug when used as a PP.
This shows that for other data-race tools,
suppressing malloc-recycle candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, \quicksand~was able to correctly identify the 69 violations of that heuristic (among 30 distinct tests),
and fall back to classifying them with DPOR.
%which we consider a much stronger verification.

{\bf Full verification.}
For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU-time.
Among these, 36 tests contained no data-race candidates whatsoever, 
so the same verification could be provided by SSS-MC.
We plot these for comparison,
according to the time taken by that test's SSS-MC control experiment,.
When a test can be totally verified without any data-race PPs,
\quicksand~may be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
Of course, using data-race PPs increases our capacity for total verification by 4.25x.

@@

\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}

Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
We classified each data-race candidate according to whether \landslide~reported them during the first interleaving,
before any backtracking or preempting:
if so, they were {\em single-pass data races}; otherwise, {\em nondeterministic}.

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{nondet-drs-1-v2.pdf}
	\caption{Some data-race candidates may not be identified during a single program execution.
		Using nondeterministic data races as PPs,
		\quicksand~found 189\% as many bugs as could be found using single-pass data races alone.
	}
	\label{fig:dr-falsenegs}
\end{figure}
Figure~\ref{fig:dr-falsenegs} compares the types of data-race candidates necessary to expose each data-race bug in our test suite.
The first series represents the bugs found using PPs from single-pass data-race candidates,
% TODO: re read racefuzzer/portend paper to confirm this is true.
i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
The second series shows all data-race bugs \quicksand~found,
which includes the former type as well as new bugs involving nondeterministic data-races.
\quicksand~found 68 data-race bugs in total, only 36 of which could be found with single-pass data-race candidates alone.
%In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.

Note that we are not comparing how much testing time is required before identifying the data-race candidates involved in each bug.
Single-pass data races can all be found after a single program execution,
while \quicksand~may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
However, prior work data-race tools \cite{tsan}, being not integrated with a MC,
are not intended to discover new candidates under subsequent runs.
Running a single-pass data-race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
\cite{chess-icb,gambit}.
Likewise, replay-based tools \cite{portend} are dependent upon the data-race detector to provide an execution trace leading to each candidate.
This experiment suggests that
such tools should be configured to use a similar feedback loop as Iterative Deepening.
%i.e., discovering transitively-reachable data races while testing initial ones.
%although that would still not simultaneously be able to provide total verifications.
% TODO: read gambit paper

@@

{\bf Finer-grained PP subsets.}
% TODO: update these numbers
\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
However, in 2 tests, no more than the minimal state space could be verified,
and in 3 other tests, no state spaces could be completed at all.
Larger state spaces often result from finer-grained locking,
which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
Such programs may require even more rigorous verification than a program with a single global lock.
%Hence these corner cases are important to consider for future work.
While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
making partial verification of such large tests possible.
%enabling partial verification of such large tests. % if need space
%Because our current implementation does not avoid repeated interleavings across state spaces,
%as discussed above,
%we were limited to a small number of very basic PP subsets to statically seed the exploration

@@

Likewise, for tests where full verification is not computationally feasible,
the state spaces associated with many data-race candidates will time out.
We cannot guarantee that those races are false positives or benign, even though no bug was found.
% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test program.
For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
which could be combined with Iterative Deepening in future work.
% Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.
